Tissue Enrichment Analysis: TEA

### 2016-03-15 tests/hgf_benchmarking/doc/tea_main_text.tex
more writing

%\documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
% \documentclass[linenumbers, doublespacing]{bmcart}
\documentclass{bmcart}

@@

%dictionaries that work well
To select between dictionaries generated between 50, 33 and 25 annotation cut-offs, and also to ensure the terms that are selected as enriched by our algorithm are reasonable, we looked in detail at the enrichment analysis results.
%Work well
Most results were highly comparable and in line with what was expected. For some sets, all dictionaries seemed to perform well. For example, in our `all neuron enriched sets'~ \cite{Spencer2011, Watson2008a} the results were an amalgamation of neuron related terms including mechanosensory neurons, thermosensitive neurons, interneurons, ganglions and male rays regardless of the dictionary used (see~\ref{fig:interagree}). On the other hand, when we looked at a gene set enriched for germline precursor expression in the embryo~ \cite{Spencer2011}, the dictionary with the 50 cutoff was only able to identify `oocyte WBbt:006797'; whereas the two smaller dictionaries were able to single out cells germline precursor cells ---at the 33-cutoff, our tool identified `Z2' and `Z3' as being five-fold enriched; whereas at the 25 gene-cutoff the terms `Psub4',`Psub3' and `Psub2' were identified in addition to `Z2' and `Z3'.
%Differences between dictionaries
We queried an embryonic stage intestine precursor geneset~ \cite{Spencer2011}. Notably, this gene set yielded no enrichment when using the 25 cutoff dictionary, nor when using the 50 cutoff dictionary. However, the 33 cutoff dictionary suggested, probably correctly, that the E lineage was heavily enriched in this set.
%Bad stuff.
Not all queries worked equally well. For example, a number of intestinal enriched genes sets~ \cite{Spencer2011, Pauli2006} were not enriched in intestine in any dictionary, but they were enriched for pharynx- and hypodermis-related terms. We were somewhat surprised that intestinal gene sets performed poorly, since the intestine is a relatively well-annotated tissue.
%Dictionary intra-agreement
We also assessed the internal agreement of our tool by using independent gene-sets that we expected to be enriched in the same tissues. We had two independent pan-neuronal sets~ \cite{Spencer2011, Watson2008a}; two independent PVD enriched sets~ \cite{Spencer2011, Smith2010}; two independent GABAergic gene sets~ \cite{Spencer2011, Cinar2005}; two independent pharyngeal gene sets~ \cite{Spencer2011, Gaudet2004a}; and two independent intestinal gene sets~ \cite{Spencer2011, Pauli2006}. Overall, the tool seems to have good internal agreement. On most sets, the same terms were enriched, although order was somewhat variable see ~\ref{fig:intragree}. However, most high-scoring terms were preserved between gene sets. The intestinal gene-sets and pharyngeal gene sets comparisons were exceptions, since at least one gene set was missing each for intestine and pharynx in every dictionary, so we didn't consider them as informative for assessing internal agreement.
 All comparisons can be found online in our Github repository (see Availability of data and materials). Overall, the dictionary generated by a 33 gene annotation cutoff with 0.95 redundancy threshold using the `any' criterion. seemed to perform well, with a good balance between specificity, verbosity and accuracy, so we selected this parameter set to generate our static dictionary.

@@

  \captionsetup{width= 0.95\textwidth}
  \caption{
  	\csentence{Schematic representation of trimming filters for an acyclical ontology.}
    \csentence{\textbf{a)}} The parent node (green) contains at least as many annotations as the union of the two sisters. These two sisters share annotations extensively. Therefore they are too similar and should be removed.
	\csentence{\textbf{b)}} Nodes with less than a threshold number of genes are trimmed (light red) and discarded from the dictionary. Here, the example threshold is 25 genes.
	\csentence{\textbf{c)}} We trim parent nodes (light red) if all their daughter nodes have more than the threshold number of annotations.
  }
\end{figure}

%   \begin{figure}[h!]
%   \caption{\csentence{Schematic diagram of annotations for two sisters.}
%       The parent node (green) contains at least as many annotations as the union of the two sisters.
% 	  These two sisters share annotations extensively. Therefore they are too similar and should be removed.
% 	  }
% 	  % \label{fig:simdiagram}
%   \end{figure}
%
% \begin{figure}[h!]
%   \caption{\csentence{Schematic showing terminal node removal.}
%       Nodes with less than a threshold number of genes are trimmed (light red) and discarded from the dictionary. Here, the threshold is 25 genes.}
% 	  \label{fig:trim_ends}
%       \end{figure}
%
% \begin{figure}[h!]
%   \caption{\csentence{Schematic showing root node removal.}
%       We trim parent nodes (light red) if all their daughter nodes have more than the threshold number of annotations. Here, the threshold is 25 genes.}
% 	  \label{fig:trim_roots}
%       \end{figure}

@@

\begin{figure}
  % \begin{subfigure}[b]{0.4\textwidth}
  %   \includegraphics[width=\textwidth]{figures/gui.png}
  %   \caption{}
  %   \label{fig:GUI}
  % \end{subfigure}
  % %
  % \begin{subfigure}[b]{0.4\textwidth}
    \includegraphics[width=\textwidth]{figures/guiresults.png}
    % \caption{}
  % \end{subfigure}
  \captionsetup{width= 0.95\textwidth}
  \caption{\csentence{Screenshot of results from the web GUI.}
  After inputting a gene-list, the user is provided with the results. 
  An HTML table is output with hyperlinks to the ontology terms. A 
  publication-ready graph is provided below, which can be saved by 
  dragging to the desktop. Finally, lists of the genes used and discarded
  for the analysis are also presented. 
      }
  \label{fig:GUIresults}
\end{figure}

@@

%pan neuronal
\begin{figure}
  \begin{subfigure}[b]{0.6\textwidth}
    \includegraphics[width=\textwidth]{WBPaper00031532_Larva_Pan_Neuronal_Enriched_WBbt_0003679_1603_33cutoff.png}
    \caption{}
    \label{fig:PanNeuronalWatson33}
  \end{subfigure}
  %
  \begin{subfigure}[b]{0.6\textwidth}
    \includegraphics[width=\textwidth]{WBPaper00031532_Larva_Pan_Neuronal_Enriched_WBbt_0003679_1603_50cutoff}
    \caption{}
    \label{fig:PanNeuronalWatson50}
  \end{subfigure}
  \captionsetup{width= 0.95\textwidth}
  \caption{\csentence{Comparison of Enrichment Results for two dictionaries for a pan-neuronal enriched gene set from ~ \cite{Watson2008a}.}
   \textbf{a)} Dictionary with cutoff: 33; threshold: 0.95; method: `any'
   \textbf{b)} Dictionary with cutoff: 50; threshold: 0.95; method: `any'
   }
   \label{fig:interagree}
\end{figure}

%agreements
\begin{figure}
  \begin{subfigure}[b]{0.6\textwidth}
    \includegraphics[width=\textwidth]{WBPaper00024970_GABAergic_neuron_specific_WBbt_0005190_247_33cutoff.png}
    \caption{}
    \label{fig:agreement1}
  \end{subfigure}
  %
  \begin{subfigure}[b]{0.6\textwidth} 	\includegraphics[width=\textwidth]{WBPaper00037950_GABAergic-motor-neurons_larva_enriched_WBbt_0005190_132_33cutoff.png}
    \caption{}
    \label{fig:agreement2}
  \end{subfigure}
  \captionsetup{width= 0.95\textwidth}
  \caption{\csentence{Independently derived genesets show similar results within a given dictionary}
  \textbf{a)} Gene set from  \cite{Watson2008a}. Dictionary with cutoff: 33; threshold: 0.95; method: `any'
  \textbf{b)} Gene set from  \cite{Spencer2011}. Dictionary with cutoff: 50; threshold: 0.95; method: `any'
   }
   \label{fig:intragree}
\end{figure}

%results Figures - engelmann et al reanalysis
%labels are missing.
\begin{figure}[h!]
    \includegraphics[width=\textwidth]{figures/DrechmeriaconiosporaEnrichment.png}
	\captionsetup{width= 0.95\textwidth}
  	\caption{\csentence{\emph{D. coniospora} enrichment results}
      Genes altered in \emph{C. elegans} after 24hr exposure to \emph{D. coniospora}, a fungus, show enrichment
	  in many neuronal tissues.}
	\label{fig:Dcon}
\end{figure}

\begin{figure}[h!]
    \includegraphics[width=\textwidth]{figures/HarposporiumspEnrichment.png}
	\captionsetup{width= 0.95\textwidth}
	\caption{\csentence{\emph{Harposporium sp.} enrichment results}
      Genes altered in \emph{C. elegans} after 24hr exposure to \emph{Harposporium sp.} a fungus, shows enrichment
	  in the posterior lateral ganglion, similarly to \emph{D. coniospora}. This particular fungus seems to affect
	  genes normally associated with neuronal precursor tissues as well. 
	  }
	\label{fig:Hsp}
\end{figure}

\begin{figure}[h!]
    \includegraphics[width=\textwidth]{figures/SerratiamarcescensEnrichment.png}
	\captionsetup{width= 0.95\textwidth}
  	\caption{ \csentence{\emph{Serratia marcescens} enrichment results}
      Genes altered in \emph{C. elegans} after 24hr exposure to \emph{Serratia marcescens}, a bacterium, shows 
	  down-regulation of chemosensory neurons notably ASER, ASEL, ASJ and ASK. The lateral ganglions are also 
	  affected. Up-regulated genes are overrepresented for terms involving the excretory duct cell and hypodermis.
	}  
	\label{fig:Smar}
\end{figure}

\begin{figure}[h!]
    \includegraphics[width=\textwidth]{figures/EnterococcusfaecalisEnrichment.png}
	\captionsetup{width= 0.95\textwidth}
  	\caption{\csentence{\emph{E. faecalis} enrichment results}
      Genes altered in \emph{C. elegans} after 24hr exposure to \emph{E. faecalis}, a bacterium, shows an unusual pattern:
	  No terms were identified as enriched in the down-regulated category. Surprisingly, up-regulated genes show enrichment
	  of various head neurons that were down-regulated with exposure to other pathogens. Interestingly, some pharyngeal
	  muscle terms are also enriched. 
	  }
	\label{fig:Efaec}
\end{figure}

\begin{figure}[h!]
    \includegraphics[width=\textwidth]{figures/OtorhabdusluminescensEnrichment.png}
	\captionsetup{width= 0.95\textwidth}
  	\caption{\csentence{\emph{O. luminescens} enrichment results}
      Genes altered in \emph{C. elegans} after 24hr exposure to \emph{O. luminescens}, a bacterium, shows no
	  significantly enriched terms in down-regulated genes. However, up-regulated genes suggest that the hypodermis
	  and excretory duct cell may be affected by infection. 
	  }
	\label{fig:Olum}
\end{figure}

>>>>

%\documentclass[twocolumn]{bmcart}% uncomment this for twocolumn layout and comment line below
\documentclass[linenumbers, doublespacing]{bmcart}
% \documentclass{bmcart}

@@

%dictionaries that work well
To select between dictionaries generated between 50, 33 and 25 annotation cut-offs, and also to ensure the terms that are selected as enriched by our algorithm are reasonable, we looked in detail at the enrichment analysis results.
%Work well
Most results were highly comparable and in line with what was expected. For some sets, all dictionaries seemed to perform well. For example, in our `all neuron enriched sets'~ \cite{Spencer2011, Watson2008a} the results were an amalgamation of neuron related terms including mechanosensory neurons, thermosensitive neurons, interneurons, ganglions and male rays regardless of the dictionary used (see~\ref{fig:interagree}). On the other hand, when we looked at a gene set enriched for germline precursor expression in the embryo~ \cite{Spencer2011}, the dictionary with the 50 cutoff was only able to identify `oocyte WBbt:006797'; whereas the two smaller dictionaries were able to single out cells germline precursor cells ---at the 33-cutoff, our tool identified `Z2' and `Z3' as being five-fold enriched; whereas at the 25 gene-cutoff the terms `Psub4',`Psub3' and `Psub2' were identified in addition to `Z2' and `Z3'.
%Differences between dictionaries
We queried an embryonic stage intestine precursor geneset~ \cite{Spencer2011}. Notably, this gene set yielded no enrichment when using the 25 cutoff dictionary, nor when using the 50 cutoff dictionary. However, the 33 cutoff dictionary suggested, probably correctly, that the E lineage was heavily enriched in this set.
%Bad stuff.
Not all queries worked equally well. For example, a number of intestinal enriched genes sets~ \cite{Spencer2011, Pauli2006} were not enriched in intestine in any dictionary, but they were enriched for pharynx- and hypodermis-related terms. We were somewhat surprised that intestinal gene sets performed poorly, since the intestine is a relatively well-annotated tissue.
%Dictionary intra-agreement
We also assessed the internal agreement of our tool by using independent gene-sets that we expected to be enriched in the same tissues. We had two independent pan-neuronal sets~ \cite{Spencer2011, Watson2008a}; two independent PVD enriched sets~ \cite{Spencer2011, Smith2010}; two independent GABAergic gene sets~ \cite{Spencer2011, Cinar2005}; two independent pharyngeal gene sets~ \cite{Spencer2011, Gaudet2004a}; and two independent intestinal gene sets~ \cite{Spencer2011, Pauli2006}. Overall, the tool seems to have good internal agreement. On most sets, the same terms were enriched, although order was somewhat variable (see Figure ~\ref{fig:intragree}). However, most high-scoring terms were preserved between gene sets. The intestinal gene-sets and pharyngeal gene sets comparisons were exceptions, since at least one gene set was missing each for intestine and pharynx in every dictionary, so we didn't consider them as informative for assessing internal agreement.
 All comparisons can be found online in our Github repository (see Availability of data and materials). Overall, the dictionary generated by a 33 gene annotation cutoff with 0.95 redundancy threshold using the `any' criterion. seemed to perform well, with a good balance between specificity, verbosity and accuracy, so we selected this parameter set to generate our static dictionary.

@@

  \captionsetup{width= 0.95\textwidth}
  \caption{
  	\csentence{Schematic representation of trimming filters for an acyclical ontology.}
    \csentence{\textbf{a)}} The parent node (green) contains at least as many annotations as the union of the two sisters. These two sisters share annotations extensively. Therefore they are too similar and should be removed.
	\csentence{\textbf{b)}} Nodes with less than a threshold number of genes are trimmed (light red) and discarded from the dictionary. Here, the example threshold is 25 genes.
	\csentence{\textbf{c)}} We trim parent nodes recursively, starting from the root, if all their daughter nodes have more than the threshold number of annotations (light red).
  }
\end{figure}

@@

\begin{figure}
	\centering
    \includegraphics[width=0.95\textwidth]{figures/guiresults.png}
  	\captionsetup{width= 0.95\textwidth}
 	\caption{\csentence{Screenshot of results from the web GUI.}
  After inputting a gene-list, the user is provided with the results. 
  An HTML table is output with hyperlinks to the ontology terms. A 
  publication-ready graph is provided below, which can be saved by 
  dragging to the desktop. Finally, lists of the genes used and discarded
  for the analysis are also presented.
  }
  \label{fig:GUIresults}
\end{figure}

@@

%pan neuronal
\begin{figure}
  \begin{subfigure}[b]{0.95\textwidth}
    \includegraphics[width=0.95\textwidth]{WBPaper00031532_Larva_Pan_Neuronal_Enriched_WBbt_0003679_1603_33cutoff.png}
    \caption{}
    \label{fig:PanNeuronalWatson33}
  \end{subfigure}
  %
  \begin{subfigure}[b]{0.95\textwidth}
    \includegraphics[width=0.95\textwidth]{WBPaper00031532_Larva_Pan_Neuronal_Enriched_WBbt_0003679_1603_50cutoff}
    \caption{}
    \label{fig:PanNeuronalWatson50}
  \end{subfigure}
  \captionsetup{width= 0.95\textwidth}
  \caption{\csentence{Comparison of Enrichment Results for two dictionaries for a pan-neuronal enriched gene set from ~\cite{Watson2008a}.}
   \textbf{a)} Dictionary with cutoff: 33; threshold: 0.95; method: `any'
   \textbf{b)} Dictionary with cutoff: 50; threshold: 0.95; method: `any'
   }
   \label{fig:interagree}
\end{figure}

%agreements
\begin{figure}
  \begin{subfigure}[b]{0.95\textwidth}
    \includegraphics[width=0.95\textwidth]{WBPaper00024970_GABAergic_neuron_specific_WBbt_0005190_247_33cutoff.png}
    \caption{}
    \label{fig:agreement1}
  \end{subfigure}
  %
  \begin{subfigure}[b]{0.95\textwidth} 	\includegraphics[width=0.95\textwidth]{WBPaper00037950_GABAergic-motor-neurons_larva_enriched_WBbt_0005190_132_33cutoff.png}
    \caption{}
    \label{fig:agreement2}
  \end{subfigure}
  \captionsetup{width= 0.95\textwidth}
  \caption{\csentence{Independently derived genesets show similar results within a given dictionary}
  \textbf{a)} Gene set from~\cite{Watson2008a}. Dictionary with cutoff: 33; threshold: 0.95; method: `any'
  \textbf{b)} Gene set from~\cite{Spencer2011}. Dictionary with cutoff: 50; threshold: 0.95; method: `any'
   }
   \label{fig:intragree}
\end{figure}

%results Figures - engelmann et al reanalysis
%labels are missing.
\begin{figure}[h!]
    \includegraphics[width=0.95\textwidth]{figures/DrechmeriaconiosporaEnrichment.png}
	\captionsetup{width= 0.95\textwidth}
  	\caption{\csentence{\emph{D. coniospora} enrichment results}
      Genes altered in \emph{C. elegans} after 24hr exposure to \emph{D. coniospora}, a fungus, show enrichment
	  in many neuronal tissues.}
	\label{fig:Dcon}
\end{figure}

\begin{figure}[h!]
    \includegraphics[width=0.95\textwidth]{figures/HarposporiumspEnrichment.png}
	\captionsetup{width= 0.95\textwidth}
	\caption{\csentence{\emph{Harposporium sp.} enrichment results}
      Genes altered in \emph{C. elegans} after 24hr exposure to \emph{Harposporium sp.} a fungus, shows enrichment
	  in the posterior lateral ganglion, similarly to \emph{D. coniospora}. This particular fungus seems to affect
	  genes normally associated with neuronal precursor tissues as well. 
	  }
	\label{fig:Hsp}
\end{figure}

\begin{figure}[h!]
    \includegraphics[width=0.95\textwidth]{figures/SerratiamarcescensEnrichment.png}
	\captionsetup{width= 0.95\textwidth}
  	\caption{ \csentence{\emph{Serratia marcescens} enrichment results}
      Genes altered in \emph{C. elegans} after 24hr exposure to \emph{Serratia marcescens}, a bacterium, shows 
	  down-regulation of chemosensory neurons notably ASER, ASEL, ASJ and ASK. The lateral ganglions are also 
	  affected. Up-regulated genes are overrepresented for terms involving the excretory duct cell and hypodermis.
	}  
	\label{fig:Smar}
\end{figure}

\begin{figure}[h!]
    \includegraphics[width=0.95\textwidth]{figures/EnterococcusfaecalisEnrichment.png}
	\captionsetup{width= 0.95\textwidth}
  	\caption{\csentence{\emph{E. faecalis} enrichment results}
      Genes altered in \emph{C. elegans} after 24hr exposure to \emph{E. faecalis}, a bacterium, shows an unusual pattern:
	  No terms were identified as enriched in the down-regulated category. Surprisingly, up-regulated genes show enrichment
	  of various head neurons that were down-regulated with exposure to other pathogens. Interestingly, some pharyngeal
	  muscle terms are also enriched. 
	  }
	\label{fig:Efaec}
\end{figure}

\begin{figure}[h!]
    \includegraphics[width=0.95\textwidth]{figures/OtorhabdusluminescensEnrichment.png}
	\captionsetup{width= 0.95\textwidth}
  	\caption{\csentence{\emph{O. luminescens} enrichment results}
      Genes altered in \emph{C. elegans} after 24hr exposure to \emph{O. luminescens}, a bacterium, shows no
	  significantly enriched terms in down-regulated genes. However, up-regulated genes suggest that the hypodermis
	  and excretory duct cell may be affected by infection. 
	  }
	\label{fig:Olum}
\end{figure}
